---
title: "Endo DEP TSST LMM"
author: "AGC"
date: "26 4 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      messages =F, 
                      warning = F)

library(haven)
library(tidyverse)
library(data.table)
library(lme4)
library(kableExtra)
library(caret)
library(ggplot2)
library(compareGroups)
library(lm.beta)

home = getwd()

```

# Prepare

## Definitions


**DV:**

* psychologischer Stress („stressed")
* Cortisol („CORT")
* Testosteron („TEST") 
* Oxytocin (OXT")
* log(TEST/CORT)

**IV of no interest :** 

* Age scaled ("age_meancentered")
* pubertal status ("pubcatimp_meancentered")
* "BMI_meancentered" #not in model * "stressed"
* "smoking_yes_no" #not in model * "stressed"


**IV of interest :**

* Gruppe ("group") = 2-Faktor HC vs DEP [Dep ist mit 3, HCs mit 2 kodiert]
* sex ("gender")
* Zeitpunkt ("Time") with 2xpolynomes
* Gruppe x Zeitpunkt(Time each poly)
* Gruppe x sex
* Gruppe x sex x Zeitpunkt(Time each poly)


**random effects**

* individual id


**sensitivity check 1**

* quantitative DEP "DIKJ_RW" [instead of group]
* gender 


**sensitivity check 2**

* ADHD lifetime diagnosis "ADHD_life"
* Essstörung "Eat_life"
* kombinierte ANX/PTSD "ANXPTSD_life"


**not included parameters**

* Medication (was included here)
* IQ
* Parental education
* Depression lifetime diagnosis
* PTSD lifetime diagnosis
* SUD lifetime diagnosis
* Anxiety lifetime diagnosis
* explstart_meancentered_min

## read and check data 

```{r preprocess, fig.height=10, fig.width=10}

# df_ANS = as.data.frame(read_sav(paste0(home,"/data/DEPHC_GU_DeskriptTSST_VUAMS_N=170 (175-5Dep).sav")))

df_ENDO = as.data.frame(read_sav(paste0(home,"/data/DEPHC_GU_DeskriptTSSThormones_175_final.sav")))

df_ENDO<-df_ENDO %>% rename(CORT_40_log=CORT_44_log)

df_ENDO$gender = drop(factor(df_ENDO$gender, levels = c(1,2), 
                        labels=c("female", "male")))%>% relevel(., ref="male")


df_ENDO$group = drop(factor(df_ENDO$group, levels = c(2,3), labels=c("HC","DEP"))) %>% relevel(., ref="HC")
df_ENDO$twuid = as.factor(df_ENDO$twuid)
df_ENDO$ADHD_life = drop(factor(df_ENDO$ADHD_life, levels = c(0,1), 
                               labels=c("no_ADHD", "ADHD")))%>% relevel(., ref="no_ADHD")

df_ENDO$Eat_life = drop(factor(df_ENDO$Eat_life, levels = c(0,1), 
                               labels=c("no_EatDis", "EatDis")))%>% relevel(., ref="no_EatDis")

df_ENDO$ANXPTSD_life = drop(factor(df_ENDO$ANXPTSD_life, levels = c(0,1), 
                               labels=c("no_ANXPTSD", "ANXPTSD")))%>% relevel(., ref="no_ANXPTSD")

df_ENDO$any_med_ccept = drop(factor(df_ENDO$any_med_ccept, levels = c(0,1), 
                               labels=c("no_med", "med")))%>% relevel(., ref="no_med")
df_ENDO$smoking_yes_no = drop(factor(df_ENDO$smoking_yes_no, levels = c(0,1), 
                                labels=c("no_smk", "smk")))%>% relevel(., ref="no_smk")


# table(df_ENDO$group, useNA = "always")
# table(df_ENDO$any_med_ccept, useNA = "always")
# table(df_ENDO$smoking_yes_no, useNA = "always")
# table(df_ENDO$gender, useNA = "always")

df_ENDO$TESTCORTlogBL = log(df_ENDO$TEST_BL_log/df_ENDO$CORT_BL_log)
df_ENDO$TESTCORTlogBL[is.infinite(df_ENDO$TESTCORTlogBL)] = NA

df_ENDO$TESTCORTlog10 = log(df_ENDO$TEST_10_log/df_ENDO$CORT_10_log)
df_ENDO$TESTCORTlog10[is.infinite(df_ENDO$TESTCORTlog10)] = NA

df_ENDO$TESTCORTlog55 = log(df_ENDO$TEST_55_log/df_ENDO$CORT_55_log)
df_ENDO$TESTCORTlog55[is.infinite(df_ENDO$TESTCORTlog55)] = NA


UV = c("age_meancentered", "BMI_meancentered", "pubcatimp_meancentered",
       #"any_med_ccept", 
       "smoking_yes_no", "gender", "group")


Sensvar = c("DIKJ_RW", 
           "ADHD_life",
           "Eat_life",
           "ANXPTSD_life")

AV = list(
AV_stressed = c(stressed_1=0, stressed_2=2,
                stressed_3=10, stressed_4=16,
                stressed_5=25, stressed_6=40, 
                stressed_7=55, stressed_8=70), 

AV_CORT = c(CORT_BL_log=0, CORT_10_log=25, CORT_25_log=40, CORT_40_log=55, CORT_55_log=70) ,
AV_TEST = c(TEST_BL_log=0, TEST_10_log=25, TEST_55_log=70),
AV_TESTCORT = c(TESTCORTlogBL = 0, TESTCORTlog10=25, TESTCORTlog55 = 70),
AV_OXT =  c(OXT_BL_log=0, OXT_1_log=16, OXT_10_log=25))

vartoplot = c(UV, Sensvar)
tmpframe = df_ENDO[,vartoplot[!vartoplot %in% c("centre")]]

tmpframe=tmpframe %>% 
  mutate_if(is.factor, function(x) as.numeric(x)-1) 

corrplot::corrplot(cor(tmpframe, use = "pairwise"))

```


## complete cohort descriptives

```{r preprocess_allstat}

vars=c(UV, unlist(lapply(AV, names)), Sensvar)

res = compareGroups(group~., data = df_ENDO[,vars])
#summary(res)
export_table <- createTable(res)
export2md(export_table)

```

## male only cohort descriptives
```{r preprocess_malestat}
res = compareGroups(group~., data = df_ENDO[,c(UV, unlist(lapply(AV, names)), Sensvar)], 
                    subset = gender=="male")
export_table <- createTable(res)
export2md(export_table)

```

## female only cohort descriptives
```{r preprocess_femalestat}
res = compareGroups(group~., 
                    data = df_ENDO[,c(UV, unlist(lapply(AV, names)), Sensvar)], 
                    subset = gender=="female")
export_table <- createTable(res)
export2md(export_table)
```


# linear model with mixed effects and time as polynomial  

We adapted a boxed design by individual over Time with a 2 polynomial decomposion of time
no site effect included as all samples where from Frankfurt 

```{r modeldef_poly}

bio.model = "DV~1+age_meancentered+BMI_meancentered+pubcatimp_meancentered+smoking_yes_no+gender+group+poly(Time, 2)+gender*group+poly(Time, 2)*group+gender*poly(Time, 2)*group+(1|twuid)"

psych.model = "DV~1+age_meancentered+pubcatimp_meancentered+gender+group+poly(Time, 2)+gender*group+poly(Time, 2)*group+gender*poly(Time, 2)*group+(1|twuid)"


bio.model.nogen = "DV~1+age_meancentered+BMI_meancentered+pubcatimp_meancentered+smoking_yes_no+group+poly(Time, 2)+poly(Time, 2)*group+(1|twuid)"

psych.model.nogen = "DV~1+age_meancentered+pubcatimp_meancentered+group+poly(Time, 2)+poly(Time, 2)*group+(1|twuid)"


bio.model.quantcd = "DV~1+age_meancentered+BMI_meancentered+pubcatimp_meancentered+smoking_yes_no+gender+DIKJ_RW+poly(Time, 2)+gender*DIKJ_RW+poly(Time, 2)*DIKJ_RW+gender*poly(Time, 2)*DIKJ_RW+(1|twuid)"

psych.model.quantcd = "DV~1+age_meancentered+pubcatimp_meancentered+gender+DIKJ_RW+poly(Time, 2)+gender*DIKJ_RW+poly(Time, 2)*DIKJ_RW+gender*poly(Time, 2)*DIKJ_RW+(1|twuid)"


models=list(
  AV_stressed = psych.model,
  AV_CORT = bio.model,
  AV_TEST = bio.model,
  AV_TESTCORT = bio.model,
  AV_OXT = bio.model)

models.nogen=list(
  AV_stressed = psych.model.nogen,
  AV_CORT = bio.model.nogen,
  AV_TEST = bio.model.nogen,
  AV_TESTCORT = bio.model.nogen,
  AV_OXT = bio.model.nogen)

models.quant=list(
  AV_stressed = psych.model.quantcd,
  AV_CORT = bio.model.quantcd,
  AV_TEST = bio.model.quantcd,
  AV_TESTCORT = bio.model.quantcd,
  AV_OXT = bio.model.quantcd)



```


```{r mml_2, warning=FALSE}

lm.beta.lmer <- function(mod) {
   b <- fixef(mod)[-1]
   sd.x <- apply(getME(mod,"X")[,-1],2,sd)
   sd.y <- sd(getME(mod,"y"))
   b*sd.x/sd.y
}

resall = list()

for (depvar in names(AV)){
  cols = names(AV[[depvar]])
  long = df_ENDO[,c("twuid",cols, UV, Sensvar)] %>% 
    gather(key = "value", value = "DV", all_of(cols))
  long$twuid = as.factor(long$twuid)
  long2  = long%>% mutate_if(is.numeric, scale)
  long2$Time = AV[[depvar]][long2$value]
  long2$value =as.factor(long2$value) %>% 
    relevel(., ref=grep("BL|stressed_1",  unique(long2$value),value = T))
  model.lme = lme4::lmer(models[[depvar]], data=long2)
  model.lme0 = lme4::lmer(DV~1+(1|twuid), data=long2)
  anovah0 = anova(model.lme0, model.lme)
  model_p_val = anovah0$`Pr(>Chisq)`[2]
  Res = summary(model.lme)
  resall[[depvar]] = model.lme
  resall[[paste0(depvar,"_longdat")]] = long2
  res.coeff = as.data.frame(Res$coefficients)
  res.coeff$stbeta = c(NA,lm.beta.lmer(model.lme))
  res.coeff$pvalue = pt(abs(res.coeff$"t value"), 1000000, lower.tail = F) * 2
  resall[[paste0(depvar,"_coeff")]]=res.coeff
  resall[[paste0(depvar,"_modsig")]]=model_p_val
  
  ## DEP quant
  model.lme = lme4::lmer(models.quant[[depvar]], data=long2)
  Res = summary(model.lme)
  resall[[depvar]] = model.lme
  resall[[paste0(depvar,"_longdat")]] = long2
  res.coeff = as.data.frame(Res$coefficients)
  res.coeff$stbeta = c(NA,lm.beta.lmer(model.lme))
  res.coeff$pvalue = pt(abs(res.coeff$"t value"), 1000000, lower.tail = F) * 2
  resall[[paste0(depvar,"_coeff_DEPquant")]]=res.coeff
  
  ## nogenmod 
  ### males
  model.lme = lme4::lmer(models.nogen[[depvar]], data=long2, subset = long2$gender=="male")
  Res = summary(model.lme)
  resall[[depvar]] = model.lme
  resall[[paste0(depvar,"_longdat")]] = long2
  res.coeff = as.data.frame(Res$coefficients)
  res.coeff$stbeta = c(NA,lm.beta.lmer(model.lme))
  res.coeff$pvalue = pt(abs(res.coeff$"t value"), 1000000, lower.tail = F) * 2
  resall[[paste0(depvar,"_coeff_males")]]=res.coeff
  
  ### females
  model.lme = lme4::lmer(models.nogen[[depvar]], data=long2, subset = long2$gender=="female")
  Res = summary(model.lme)
  resall[[depvar]] = model.lme
  resall[[paste0(depvar,"_longdat")]] = long2
  res.coeff = as.data.frame(Res$coefficients)
  res.coeff$stbeta = c(NA,lm.beta.lmer(model.lme))
  res.coeff$pvalue = pt(abs(res.coeff$"t value"), 1000000, lower.tail = F) * 2
  resall[[paste0(depvar,"_coeff_females")]]=res.coeff

  for (sv in Sensvar){
      model.lme = lme4::lmer(str_replace(models[[depvar]], "~1+",paste0("~1+",sv,"+")), data=long2)
      Res = summary(model.lme)
      res.coeff = as.data.frame(Res$coefficients)
      res.coeff$stbeta = c(NA,lm.beta.lmer(model.lme))
      res.coeff$pvalue = pt(abs(res.coeff$"t value"), 1000000, lower.tail = F) * 2
      resall[[paste0(depvar,"_coeff", "_",sv)]]=res.coeff
  }
}


```

## Results 

### stressed

full models: `r models[[1]]`

h0 model: DV~1+(1|twuid):

overall model p-value:`r as.character(signif(resall[["AV_stressed_modsig"]],3))`

```{r, wrap_res_stressed_2}
tableplot = function (x){
  x %>% dplyr::mutate_if(is.numeric, function(x){as.character(signif(x, 3))}) %>% kbl() %>% kable_classic()
  }


depvar = "AV_stressed"
labeltag = "Psychological stress"

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(DV, group=Time, col=Time)) + 
  ylab("density") + xlab(labeltag)+
  geom_density()

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(Time,DV, col=group)) + 
  ylab(labeltag) + xlab("Time")+
  geom_smooth(method = 'loess') + geom_point() +   facet_wrap(~gender)
```

#### sensitivity analyses

```{r, sensitivity_stressed}

resall[[paste0(depvar, "_coeff")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_DEPquant")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_males")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_females")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ADHD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_Eat_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ANXPTSD_life")]] %>% tableplot()

```

### CORT


full models: `r models[[2]]`

h0 model:DV~1+(1|twuid):

overall model p-value:`r as.character(signif(resall[["AV_CORT_modsig"]],3))`


```{r wrap_res_CORT_2}
depvar = "AV_CORT"
labeltag = "Cortisol"

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(DV, group=Time, col=Time)) + 
  ylab("density") + xlab(labeltag)+
  geom_density()

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(Time,DV, col=group)) + 
  ylab(labeltag) + xlab("Time")+
  geom_smooth(method = 'loess') + geom_point() +   facet_wrap(~gender)

```

#### sensitivity analyses

```{r, sensitivity_cort}

resall[[paste0(depvar, "_coeff")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_DEPquant")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_males")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_females")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ADHD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_Eat_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ANXPTSD_life")]] %>% tableplot()


```

### TEST


full models: `r models[[3]]`

h0 model: DV~1+(1|twuid):


overall model p-value:
`r as.character(signif(resall[["AV_TEST_modsig"]], 3))`


```{r wrap_res_TEST_2}
depvar = "AV_TEST"
labeltag = "Testosterone"

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(DV, group=Time, col=Time)) + 
  ylab("density") + xlab(labeltag)+
  geom_density()

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(Time,DV, col=group)) + 
  ylab(labeltag) + xlab("Time")+
  geom_smooth(method = 'loess') + geom_point() +   facet_wrap(~gender)


```

#### sensitivity analyses

```{r, sensitivity_test}

resall[[paste0(depvar, "_coeff")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_DEPquant")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_males")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_females")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ADHD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_Eat_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ANXPTSD_life")]] %>% tableplot()


```

### TEST/CORT ratio


full models: `r models[[4]]`

h0 model: DV~1+(1|twuid):


overall model p-value:
`r as.character(signif(resall[["AV_TESTCORT_modsig"]], 3))`

```{r, wrap_res_testcort_2}

depvar = "AV_TESTCORT"
labeltag = "Testosterone/Cortisol ratio"

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(DV, group=Time, col=Time)) + 
  ylab("density") + xlab(labeltag)+
  geom_density()

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(Time,DV, col=group)) + 
  ylab(labeltag) + xlab("Time")+
  geom_smooth(method = 'loess') + geom_point() +   facet_wrap(~gender)

```

#### sensitivity analyses

```{r, sensitivity_testcort}

resall[[paste0(depvar, "_coeff")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_DEPquant")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_males")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_females")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ADHD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_Eat_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ANXPTSD_life")]] %>% tableplot()

```


### OXT

full models: `r models[[2]]`

h0 model: DV~DV~1+(1|twuid):

overall model p-value:
`r as.character(signif(resall[["AV_OXT_modsig"]], 3))`


```{r wrap_res_OXT_2}

depvar = "AV_OXT"
labeltag = "Oxytocin"

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(DV, group=Time, col=Time)) + 
  ylab("density") + xlab(labeltag)+
  geom_density()

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(Time,DV, col=group)) + 
  ylab(labeltag) + xlab("Time")+
  geom_smooth(method = 'loess') + geom_point() +   facet_wrap(~gender)


```

#### sensitivity analyses

```{r, sensitivity_OXT}

resall[[paste0(depvar, "_coeff")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_DEPquant")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_males")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_females")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ADHD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_Eat_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ANXPTSD_life")]] %>% tableplot()


```

```{r summarytablefunc}

# with full pvalues 
tableplot_2= function (x){
    x %>% dplyr::mutate_if(is.numeric, function(x){as.character(signif(x, 3))}) %>% 
    kbl(.,col.names = newnames) %>% 
    add_header_above(c("indep. variable" = 1, 
                       "Psych. stress"=4, 
                       "Testosterone"=4, 
                       "Cortisol"=4, 
                       "Test/Cort ratio"=4,
                       "Oxytocin"=4)) %>% kable_classic()
}

# with symblic p-values 
tableplot_3= function (x){
    x %>% dplyr::mutate_if(is.numeric, 
                           function(x){as.character(round(x, 2))}) %>%
    mutate_all(~replace(., .=="0", "<0.01")) %>%
    kbl(.,col.names = newnames) %>% 
    add_header_above(c("indep. variable" = 1, 
                       "Psych. stress"=4, 
                       "Cortisol"=4, 
                       "Testosterone"=4, 
                       "Test/Cort ratio"=4,
                       "Oxytocin"=4)) %>% kable_classic()
}


```
## summary table main models 

### all effects
```{r summarytableshow}

restabnames = grep("coeff$",names(resall), value=T)
resall_mod=lapply(resall[restabnames], tibble::rownames_to_column)
fullres=resall_mod %>% reduce(full_join, by="rowname")
fullres = fullres %>% select (.,- grep("t value", colnames(fullres), value = T))
newnames = c("-", rep(c("beta", "se", "st.beta", "P"), length(names(AV))))

#fullres %>% tableplot_2() 
fullres %>% tableplot_3() 


```

### effects of interest

```{r summarytable reduced}

index = complete.cases(fullres) & !fullres$rowname %in% c("(Intercept)","age_meancentered")

fullres[index,] %>% as.tibble() %>% tableplot_3()

```


### females 
```{r summarytableshow_females}

restabnames = grep("coeff_females$",names(resall), value=T)
resall_mod=lapply(resall[restabnames], tibble::rownames_to_column)
fullres=resall_mod %>% reduce(full_join, by="rowname")
fullres = fullres %>% select (.,- grep("t value", colnames(fullres), value = T))
newnames = c("-", rep(c("beta", "se", "st.beta", "P"), length(names(AV))))

#fullres %>% tableplot_2() 
fullres %>% tableplot_3() 

```



### males 
```{r summarytableshow_males}

restabnames = grep("coeff_males$",names(resall), value=T)
resall_mod=lapply(resall[restabnames], tibble::rownames_to_column)
fullres=resall_mod %>% reduce(full_join, by="rowname")
fullres = fullres %>% select (.,- grep("t value", colnames(fullres), value = T))
newnames = c("-", rep(c("beta", "se", "st.beta", "P"), length(names(AV))))

#fullres %>% tableplot_2() 
fullres %>% tableplot_3() 

```


### quantitative DEP
```{r summarytableshow_CDquant}

restabnames = grep("coeff_DEPquant$",names(resall), value=T)
resall_mod=lapply(resall[restabnames], tibble::rownames_to_column)
fullres=resall_mod %>% reduce(full_join, by="rowname")
fullres = fullres %>% select (.,- grep("t value", colnames(fullres), value = T))
newnames = c("-", rep(c("beta", "se", "st.beta", "P"), length(names(AV))))

#fullres %>% tableplot_2() 
fullres %>% tableplot_3() 

```
