---
title: "ANS DEP TSST LMM"
author: "AGC"
date: "26 4 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      messages =F, 
                      warning = F)

library(haven)
library(tidyverse)
library(data.table)
library(lme4)
library(kableExtra)
library(caret)
library(ggplot2)
library(compareGroups)
library(lm.beta)

home = getwd()

```


# Prepare

## Definitions


**DV:**

* psychologischer Stress („stressed")
* Average_HR_bpm
* RSA_*_trans
* PEP_*_trans # inclusion PEP_6_trans >1 all other >2 # 

**IV of no interest :** 

* Age scaled ("age_meancentered")
* pubertal status ("pubcatimp")
* "BMI_meancentered" #not in model * "stressed"
* "smoking_yes_no" #not in model * "stressed"
* "sports_meancentered"  #not in model * "stressed"
* Respiration_*_trans #only in model RSA


**IV of interest :**

* Gruppe ("group") = 2-Faktor HC vs DEP [Dep ist mit 3, HCs mit 2 kodiert]
* sex ("gender")
* Zeitpunkt ("Time") with 2xpolynomes
* Gruppe x Zeitpunkt(Time each poly)
* Gruppe x sex
* Gruppe x sex x Zeitpunkt(Time each poly)


**random effects**

* individual id


**sensitivity check 1**

* quantitative DEP "DIKJ_RW" [instead of group]
* gender 


**sensitivity check 2**

* ADHD lifetime diagnosis "ADHD_life"
* Essstörung "Eat_life"
* kombinierte ANX/PTSD "ANXPTSD_life"


**not included parameters**

* Medication (was included here)
* IQ
* Parental education
* Depression lifetime diagnosis
* PTSD lifetime diagnosis
* SUD lifetime diagnosis
* Anxiety lifetime diagnosis
* explstart_meancentered_min

## read and check data 


```{r preprocess, fig.height=10, fig.width=10}

df_ANS = as.data.frame(read_sav(paste0(home,"/data/DEPHC_GU_DeskriptTSST_VUAMS_N=170 (175-5Dep).sav")))


df_ANS$gender = drop(factor(df_ANS$gender, levels = c(1,2), 
                        labels=c("female", "male")))%>% relevel(., ref="male")


df_ANS$group = drop(factor(df_ANS$group, levels = c(2,3), labels=c("HC","DEP"))) %>% relevel(., ref="HC")
df_ANS$group_oANXPTSD_life = drop(factor(df_ANS$group_oANXPTSD_life, levels = c(2,3), labels=c("HC","DEP"))) %>% relevel(., ref="HC")
df_ANS$group_mANXPTSD_life = drop(factor(df_ANS$group_mANXPTSD_life, levels = c(2,3), labels=c("HC","DEP"))) %>% relevel(., ref="HC")


df_ANS$twuid = as.factor(df_ANS$twuid)
df_ANS$ADHD_life = drop(factor(df_ANS$ADHD_life, levels = c(0,1), 
                               labels=c("no_ADHD", "ADHD")))%>% relevel(., ref="no_ADHD")

df_ANS$Eat_life = drop(factor(df_ANS$Eat_life, levels = c(0,1), 
                               labels=c("no_EatDis", "EatDis")))%>% relevel(., ref="no_EatDis")

df_ANS$ANXPTSD_life = drop(factor(df_ANS$ANXPTSD_life, levels = c(0,1), 
                               labels=c("no_ANXPTSD", "ANXPTSD")))%>% relevel(., ref="no_ANXPTSD")


df_ANS$smoking_yes_no = drop(factor(df_ANS$smoking_yes_no, levels = c(0,1), 
                                labels=c("no_smk", "smk")))%>% relevel(., ref="no_smk")


UV = c("age_meancentered", "BMI_meancentered", "pubcatimp",
       "smoking_yes_no", "gender", "group", "sports_meancentered")


Respiration_Rate = c(Respiration_Rate_bpm.2=1, 
          Respiration_Rate_bpm.3=2, 
          Respiration_Rate_bpm.4=3, 
          Respiration_Rate_bpm.5=4, 
          Respiration_Rate_bpm.6=5, 
          Respiration_Rate_bpm.7=6) 




Sensvar = c("DIKJ_RW", #replaces group
            "group_mANXPTSD_life", #replaces group
            "group_oANXPTSD_life", #replaces group
           "ADHD_life",
           "Eat_life",
           "ANXPTSD_life")


AV = list(
AV_stressed = c(stressed_1=0, stressed_2=2,
                stressed_3=10, stressed_4=16,
                stressed_5=25, stressed_6=40, 
                stressed_7=55, stressed_8=70), 

AV_HR = c(Average_HR_bpm.2=1, 
          Average_HR_bpm.3=2, 
          Average_HR_bpm.4=3, 
          Average_HR_bpm.5=4, 
          Average_HR_bpm.6=5, 
          Average_HR_bpm.7=6), 

AV_RSA = c(RSA_2_trans=1, 
          RSA_3_trans=2, 
          RSA_4_trans=3, 
          RSA_5_trans=4, 
          RSA_6_trans=5, 
          RSA_7_trans=6), 

AV_PEP = c(PEP_2_trans=1, 
          PEP_3_trans=2, 
          PEP_4_trans=3, 
          PEP_5_trans=4, 
          PEP_6_trans=5, 
          PEP_7_trans=6))


#filter PEP
attach(df_ANS)
df_ANS$PEP_2_trans[PEP_2_trans <= 2]=NA 
df_ANS$PEP_3_trans[PEP_3_trans <= 2]=NA 
df_ANS$PEP_4_trans[PEP_4_trans <= 2]=NA 
df_ANS$PEP_5_trans[PEP_5_trans <= 2]=NA 
df_ANS$PEP_6_trans[PEP_6_trans <= 1]=NA 
df_ANS$PEP_7_trans[PEP_7_trans <= 2]=NA 
detach(df_ANS)

vartoplot = c(UV, Sensvar, names(Respiration_Rate))
tmpframe = df_ANS[,vartoplot[!vartoplot %in% c("centre")]] %>% gather(key = "value", value = "RespirationRate", all_of(names(Respiration_Rate)))

tmpframe=tmpframe %>% 
  mutate_if(is.factor, function(x) as.numeric(x)-1) %>% select(-value)

corrplot::corrplot(cor(tmpframe, use = "pairwise"))

```

## complete cohort descriptives

```{r preprocess_allstat}

vars=c(UV, names(Respiration_Rate), unlist(lapply(AV, names)), Sensvar)

res = compareGroups(group~., data = df_ANS[,vars])
#summary(res)
export_table <- createTable(res)
export2md(export_table)

```

## male only cohort descriptives
```{r preprocess_malestat}
res = compareGroups(group~., data = df_ANS[,c(UV, unlist(lapply(AV, names)), Sensvar)], 
                    subset = gender=="male")
export_table <- createTable(res)
export2md(export_table)

```


## female only cohort descriptives
```{r preprocess_femalestat}
res = compareGroups(group~., 
                    data = df_ANS[,c(UV, unlist(lapply(AV, names)), Sensvar)], 
                    subset = gender=="female")
export_table <- createTable(res)
export2md(export_table)
```



# linear model with mixed effects and time as polynomial  

We adapted a boxed design by individual over Time with a 2 polynomial decomposion of time
no site effect included as all samples where from Frankfurt 


```{r modeldef_poly}

varint = "gender+group+poly(Time, 2)+gender*group+poly(Time, 2)*group+gender*poly(Time, 2)*group+(1|twuid)"
varintq = "gender+DIKJ_RW+poly(Time, 2)+gender*DIKJ_RW+poly(Time, 2)*DIKJ_RW+gender*poly(Time, 2)*DIKJ_RW+(1|twuid)"
varint_oAnxPTSD = "gender+group_oANXPTSD_life+poly(Time, 2)+gender*group_oANXPTSD_life+poly(Time, 2)*group_oANXPTSD_life+gender*poly(Time, 2)*group_oANXPTSD_life+(1|twuid)"
varint_mAnxPTSD = "gender+group_mANXPTSD_life+poly(Time, 2)+gender*group_mANXPTSD_life+poly(Time, 2)*group_mANXPTSD_life+gender*poly(Time, 2)*group_mANXPTSD_life+(1|twuid)"

varint_nogen="group+poly(Time, 2)+poly(Time, 2)*group+(1|twuid)"
varint_nogenq="DIKJ_RW+poly(Time, 2)+poly(Time, 2)*DIKJ_RW+(1|twuid)"
varint_nogen_oAnxPTSD ="oANXPTSD+poly(Time, 2)+poly(Time, 2)*oANXPTSD+(1|twuid)"
varint_nogen_mAnxPTSD ="mANXPTSD+poly(Time, 2)+poly(Time, 2)*mANXPTSD+(1|twuid)" 


models=list(
  AV_stressed = paste0("DV~1+age_meancentered+pubcatimp+",varint),
  AV_HR=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+",varint),
  AV_RSA=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+RespirationRate+",varint),
  AV_PEP=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+",varint))

models.nogen=list(
  AV_stressed = paste0("DV~1+age_meancentered+pubcatimp+",varint_nogen),
  AV_HR=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+",varint_nogen),
  AV_RSA=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+RespirationRate+",varint_nogen),
  AV_PEP=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+",varint_nogen))

models.quant=list(
  AV_stressed = paste0("DV~1+age_meancentered+pubcatimp+",varintq),
  AV_HR=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+",varintq),
  AV_RSA=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+RespirationRate+",varintq),
  AV_PEP=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+",varintq))

models.oAnxPTSD=list(
  AV_stressed = paste0("DV~1+age_meancentered+pubcatimp+",varint_oAnxPTSD),
  AV_HR=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+",varint_oAnxPTSD),
  AV_RSA=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+RespirationRate+",varint_oAnxPTSD),
  AV_PEP=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+",varint_oAnxPTSD))

models.mAnxPTSD=list(
  AV_stressed = paste0("DV~1+age_meancentered+pubcatimp+",varint_mAnxPTSD),
  AV_HR=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+",varint_mAnxPTSD),
  AV_RSA=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+RespirationRate+",varint_mAnxPTSD),
  AV_PEP=paste0("DV~1+age_meancentered+pubcatimp+BMI_meancentered+smoking_yes_no+sports_meancentered+",varint_mAnxPTSD))


```


```{r mml_2, warning=FALSE}

lm.beta.lmer <- function(mod) {
   b <- fixef(mod)[-1]
   sd.x <- apply(getME(mod,"X")[,-1],2,sd)
   sd.y <- sd(getME(mod,"y"))
   b*sd.x/sd.y
}

resall = list()

for (depvar in names(AV)){
  cols =  names(AV[[depvar]])
  long = df_ANS[,c("twuid",cols, UV, Sensvar)] %>% 
    gather(key = "value", value = "DV", all_of(cols))
  if(depvar =="AV_RSA"){ 
    long$RespirationRate = tmpframe$RespirationRate}
  long$twuid = as.factor(long$twuid)
  long2  = long%>% mutate_if(is.numeric, scale)
  long2$Time = AV[[depvar]][long2$value]
  long2$value =as.factor(long2$value) %>% 
    relevel(., ref=grep("\\.2|_2_|stressed_1",  unique(long2$value),value = T))
  model.lme = lme4::lmer(models[[depvar]], data=long2)
  model.lme0 = lme4::lmer(DV~1+(1|twuid), data=long2)
  anovah0 = anova(model.lme0, model.lme)
  model_p_val = anovah0$`Pr(>Chisq)`[2]
  Res = summary(model.lme)
  resall[[depvar]] = model.lme
  resall[[paste0(depvar,"_longdat")]] = long2
  res.coeff = as.data.frame(Res$coefficients)
  res.coeff$stbeta = c(NA,lm.beta.lmer(model.lme))
  res.coeff$pvalue = pt(abs(res.coeff$"t value"), 1000000, lower.tail = F) * 2
  resall[[paste0(depvar,"_coeff")]]=res.coeff
  resall[[paste0(depvar,"_modsig")]]=model_p_val
  
  ## DEP quant
  model.lme = lme4::lmer(models.quant[[depvar]], data=long2)
  Res = summary(model.lme)
  resall[[depvar]] = model.lme
  resall[[paste0(depvar,"_longdat")]] = long2
  res.coeff = as.data.frame(Res$coefficients)
  res.coeff$stbeta = c(NA,lm.beta.lmer(model.lme))
  res.coeff$pvalue = pt(abs(res.coeff$"t value"), 1000000, lower.tail = F) * 2
  resall[[paste0(depvar,"_coeff_DEPquant")]]=res.coeff
  
  ## oAnxPTSD
  model.lme = lme4::lmer(models.oAnxPTSD[[depvar]], data=long2)
  Res = summary(model.lme)
  resall[[depvar]] = model.lme
  resall[[paste0(depvar,"_longdat")]] = long2
  res.coeff = as.data.frame(Res$coefficients)
  res.coeff$stbeta = c(NA,lm.beta.lmer(model.lme))
  res.coeff$pvalue = pt(abs(res.coeff$"t value"), 1000000, lower.tail = F) * 2
  resall[[paste0(depvar,"_coeff_oAnxPTSD")]]=res.coeff
  
  
  ## mAnxPTSD
  model.lme = lme4::lmer(models.mAnxPTSD[[depvar]], data=long2)
  Res = summary(model.lme)
  resall[[depvar]] = model.lme
  resall[[paste0(depvar,"_longdat")]] = long2
  res.coeff = as.data.frame(Res$coefficients)
  res.coeff$stbeta = c(NA,lm.beta.lmer(model.lme))
  res.coeff$pvalue = pt(abs(res.coeff$"t value"), 1000000, lower.tail = F) * 2
  resall[[paste0(depvar,"_coeff_oAnxPTSD")]]=res.coeff
  
  ## nogenmod 
  ### males
  model.lme = lme4::lmer(models.nogen[[depvar]], data=long2, subset = long2$gender=="male")
  Res = summary(model.lme)
  resall[[depvar]] = model.lme
  resall[[paste0(depvar,"_longdat")]] = long2
  res.coeff = as.data.frame(Res$coefficients)
  res.coeff$stbeta = c(NA,lm.beta.lmer(model.lme))
  res.coeff$pvalue = pt(abs(res.coeff$"t value"), 1000000, lower.tail = F) * 2
  resall[[paste0(depvar,"_coeff_males")]]=res.coeff
  
  ### females
  model.lme = lme4::lmer(models.nogen[[depvar]], data=long2, subset = long2$gender=="female")
  Res = summary(model.lme)
  resall[[depvar]] = model.lme
  resall[[paste0(depvar,"_longdat")]] = long2
  res.coeff = as.data.frame(Res$coefficients)
  res.coeff$stbeta = c(NA,lm.beta.lmer(model.lme))
  res.coeff$pvalue = pt(abs(res.coeff$"t value"), 1000000, lower.tail = F) * 2
  resall[[paste0(depvar,"_coeff_females")]]=res.coeff

  for (sv in Sensvar){
      model.lme = lme4::lmer(str_replace(models[[depvar]], "~1+",paste0("~1+",sv,"+")), data=long2)
      Res = summary(model.lme)
      res.coeff = as.data.frame(Res$coefficients)
      res.coeff$stbeta = c(NA,lm.beta.lmer(model.lme))
      res.coeff$pvalue = pt(abs(res.coeff$"t value"), 1000000, lower.tail = F) * 2
      resall[[paste0(depvar,"_coeff", "_",sv)]]=res.coeff
  }
}

tableplot = function (x){
  x %>% dplyr::mutate_if(is.numeric, function(x){as.character(signif(x, 3))}) %>% kbl() %>% kable_classic()
  }


```

## Results 

### stressed

full models: `r models[[1]]`

h0 model: DV~1+(1|twuid):

overall model p-value:`r as.character(signif(resall[["AV_stressed_modsig"]],3))`

```{r, wrap_res_stressed_2}

depvar = "AV_stressed"
labeltag = "Psychological stress"

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(DV, group=Time, col=Time)) + 
  ylab("density") + xlab(labeltag)+
  geom_density()

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(Time,DV, col=group)) + 
  ylab(labeltag) + xlab("Time")+
  geom_smooth(method = 'loess') + geom_point() +   facet_wrap(~gender)

```

```{r, sensitivity_stressed}
resall[[paste0(depvar, "_coeff")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_DEPquant")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_group_mANXPTSD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_group_oANXPTSD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_males")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_females")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ADHD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_Eat_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ANXPTSD_life")]] %>% tableplot()

```

### Heartrate

full models: `r models[[2]]`

h0 model: DV~1+(1|twuid):

overall model p-value:`r as.character(signif(resall[["AV_HR_modsig"]],3))`

```{r, wrap_res_HR_2}

depvar = "AV_HR"
labeltag = "Heartrate"

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(DV, group=Time, col=Time)) + 
  ylab("density") + xlab(labeltag)+
  geom_density()

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(Time,DV, col=group)) + 
  ylab(labeltag) + xlab("Time")+
  geom_smooth(method = 'loess') + geom_point() +   facet_wrap(~gender)

```

```{r, sensitivity_HR}
resall[[paste0(depvar, "_coeff")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_DEPquant")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_group_mANXPTSD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_group_oANXPTSD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_males")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_females")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ADHD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_Eat_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ANXPTSD_life")]] %>% tableplot()

```

### Respiratory Sinus Arrhythmia 

full models: `r models[[3]]`

h0 model: DV~1+(1|twuid):

overall model p-value:`r as.character(signif(resall[["AV_RSA_modsig"]],3))`

```{r, wrap_res_RSA_2}

depvar = "AV_RSA"
labeltag = "Respiratory Sinus Arrhythmia"

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(DV, group=Time, col=Time)) + 
  ylab("density") + xlab(labeltag)+
  geom_density()

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(Time,DV, col=group)) + 
  ylab(labeltag) + xlab("Time")+
  geom_smooth(method = 'loess') + geom_point() +   facet_wrap(~gender)

```

```{r, sensitivity_RSA}
resall[[paste0(depvar, "_coeff")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_DEPquant")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_group_mANXPTSD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_group_oANXPTSD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_males")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_females")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ADHD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_Eat_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ANXPTSD_life")]] %>% tableplot()

```


### Pre-ejection period 

full models: `r models[[2]]`

h0 model: DV~1+(1|twuid):

overall model p-value:`r as.character(signif(resall[["AV_PEP_modsig"]],3))`

```{r, wrap_res_PEP_2}

depvar = "AV_PEP"
labeltag = "pre-ejection period "

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(DV, group=Time, col=Time)) + 
  ylab("density") + xlab(labeltag)+
  geom_density()

ggplot(data = resall[[paste0(depvar, "_longdat")]], 
       aes(Time,DV, col=group)) + 
  ylab(labeltag) + xlab("Time")+
  geom_smooth(method = 'loess') + geom_point() +   facet_wrap(~gender)

```

```{r, sensitivity_PEP}
resall[[paste0(depvar, "_coeff")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_DEPquant")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_group_mANXPTSD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_group_oANXPTSD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_males")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_females")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ADHD_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_Eat_life")]] %>% tableplot()
resall[[paste0(depvar, "_coeff_ANXPTSD_life")]] %>% tableplot()

```

